Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-28 manuscript/manuscript.Rnw
move PRLR and HOXA11 tree to supplemnetal

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \mouse are shown in
\cref{tab:top50_mouse_placenta} with a list of all genes with
$\mathrm{FPKM} \ge 50$ in Supplemental
\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
target \textit{in silico} is the orphan G-coupled receptor
\textit{Mrgprg}
\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
of which are maternally imprinted (the later of which is a homolog of
human \textit{H19\_2}),
\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
cathepsins and a rodentia specific expansion of prolactin related
genes. There are also two miRNA which overlap with the of the 7S
signal recognition particle
(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
and
\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
and numerous pseudogenes
(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
appear to be highly expressed, however, the short read length (34--44)
in many samples is likely preventing unambiguous assignment of all
reads to the more lengthy complete transcript, so these shorter
transcripts should be treated with suspicion, and we will ignore them
in further discussion. The prolactin-related genes
(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
very highly expressed in \mouse in, with %
\Sexpr{combined.fpkm[species=="mus musculus" &
  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
prolactin related genes having $\mathrm{FPKM} \ge 1$
(\cref{fig:prolactin_evolution_tree}). The cathespins have also
expanded in rodents, with six showing high expression
(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
\cref{tab:top50_mouse_placenta,fig:cathespin_evolution_tree}).
Finally, \textit{Peg10}, an imprinted gene required for proper
formation of the placenta in mice is also highly expressed
(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.

@@

The elephant is strikingly the only mammal analyzed that shows high
expression of the primarily pituitary hormone prolactin (\textit{PRL})
in the placenta, although we see high expression of genes which
resulted from duplications of \textit{PRL} in the \mouse, \carmeli,
and \cow (\cref{fig:prolactin_evolution_tree}). All afrotherians yet
analyzed contain only one prolactin gene, and prolactin has undergone
adaptive evolution on the \elephant lineage
\cite{Wallis2009:ProlactininAfrotheriacharacterization}.
\citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
presence of a prolactin in the trophoblast of the elephant placenta
using immunohistological staining and human derived antibodies. A
previous analysis of \elephant placenta also detected the expression
of prolactin \cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
Here we confirm that the sole prolactin in \elephant is not only
expressed but highly expressed in the placenta, and likely functions
both in the maintenance of the corpus luteum and as a placental
lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}.
\textit{PRL} is not expressed in \opossum, even though there is
evidence that it is expressed in the placenta of another marsupial,
\textit{Macropus eugenii} (tammar wallaby)
\cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests
that placental prolactin expression in marsupials may be as diverse as
in eutherian mammals. One possible explanation of the diversity of
expression is the presence of endogenous retroviruses such as MER39 in
primates, MER77 in rodents, and L1-2\_LA in \elephant
\cite{Emera.ea2012:Convergentevolutionofendometrial} although this
does not explain the diversity of expression of different isoforms of
\textit{PRL} in \mouse or their tissue specificity in primates. An
alternative, that \textit{HOXA11} expression is responsible for
\textit{PRL} expression
\cite{Lynch.ea2008:Adaptivechangesintranscription} seems not to be the
case in placenta, as the rodent with the highest \textit{HOXA11}
expression, \galili
(\Sexpr{gene.expression("nannospalax_galili","Hoxa11",complete=TRUE)})
is the rodent with the lowest expression of \textit{PRL}
(\Sexpr{gene.expression("nannospalax_galili","Prl",complete=TRUE)},
\cref{fig:prolactin_evolution_tree}), and \elephant also does not
express \textit{HOXA11} highly either
(\Sexpr{gene.expression("loxodonta_africana","HOXA11",complete=TRUE)}).

@@

\begin{figure}
<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRLR" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 subtree=c("ENSG00000113494","ENSMODG00000020376"))
@ 
\caption{\genetreeexplanation{prolactin receptor}}
\label{fig:prolactin_receptor_evolution_tree}
\end{figure}

\begin{figure}
<<hoxa11.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="HOXA11" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 subtree=c("ENSBTAG00000014738","ENSECAG00000013229")
                 )
@ 
\caption{\genetreeexplanation{\textit{HOXA11}}}
\label{fig:hoxa11_evolution_tree}
\end{figure}

>>>>

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \mouse are shown in
\cref{tab:top50_mouse_placenta} with a list of all genes with
$\mathrm{FPKM} \ge 50$ in Supplemental
\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
target \textit{in silico} is the orphan G-coupled receptor
\textit{Mrgprg}
\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
of which are maternally imprinted (the later of which is a homolog of
human \textit{H19\_2}),
\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
cathepsins and a rodentia specific expansion of prolactin related
genes. There are also two miRNA which overlap with the of the 7S
signal recognition particle
(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
and
\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
and numerous pseudogenes
(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
appear to be highly expressed, however, the short read length (34--44)
in many samples is likely preventing unambiguous assignment of all
reads to the more lengthy complete transcript, so these shorter
transcripts should be treated with suspicion, and we will ignore them
in further discussion. The prolactin-related genes
(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
very highly expressed in \mouse in, with %
\Sexpr{combined.fpkm[species=="mus musculus" &
  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
prolactin related genes having $\mathrm{FPKM} \ge 1$
(\cref{fig:prolactin_evolution_tree}, Supplemental
\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
also expanded in rodents, with six showing high expression
(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
\cref{tab:top50_mouse_placenta,fig:cathespin_evolution_tree}).
Finally, \textit{Peg10}, an imprinted gene required for proper
formation of the placenta in mice is also highly expressed
(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.

@@

The elephant is strikingly the only mammal analyzed that shows high
expression of the primarily pituitary hormone prolactin (\textit{PRL})
in the placenta, although we see high expression of genes which
resulted from duplications of \textit{PRL} in the \mouse, \carmeli,
and \cow (\cref{fig:prolactin_evolution_tree}). All afrotherians yet
analyzed contain only one prolactin gene, and prolactin has undergone
adaptive evolution on the \elephant lineage
\cite{Wallis2009:ProlactininAfrotheriacharacterization}.
\citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
presence of a prolactin in the trophoblast of the elephant placenta
using immunohistological staining and human derived antibodies. A
previous analysis of \elephant placenta also detected the expression
of prolactin \cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
Here we confirm that the sole prolactin in \elephant is not only
expressed but highly expressed in the placenta, and likely functions
both in the maintenance of the corpus luteum and as a placental
lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}.
\textit{PRL} is not expressed in \opossum, even though there is
evidence that it is expressed in the placenta of another marsupial,
\textit{Macropus eugenii} (tammar wallaby)
\cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests
that placental prolactin expression in marsupials may be as diverse as
in eutherian mammals. One possible explanation of the diversity of
expression is the presence of endogenous retroviruses such as MER39 in
primates, MER77 in rodents, and L1-2\_LA in \elephant
\cite{Emera.ea2012:Convergentevolutionofendometrial} although this
does not explain the diversity of expression of different isoforms of
\textit{PRL} in \mouse or their tissue specificity in primates. An
alternative, that \textit{HOXA11} expression is responsible for
\textit{PRL} expression
\cite{Lynch.ea2008:Adaptivechangesintranscription} seems not to be the
case in placenta, as the rodent with the highest \textit{HOXA11}
expression, \galili
(\Sexpr{gene.expression("nannospalax_galili","Hoxa11",complete=TRUE)})
is the rodent with the lowest expression of \textit{PRL}
(\Sexpr{gene.expression("nannospalax_galili","Prl",complete=TRUE)},
\cref{fig:prolactin_evolution_tree}), and \elephant also does not
express \textit{HOXA11} highly either
(\Sexpr{gene.expression("loxodonta_africana","HOXA11",complete=TRUE)},
Supplemental \cref{fig:hoxa11_evolution_tree}).

@@

### 2016-01-28 manuscript/supplemental_material.Rnw
move PRLR and HOXA11 tree to supplemnetal

>>>>

\begin{figure}
<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRLR" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 subtree=c("ENSG00000113494","ENSMODG00000020376"))
@ 
\caption{\genetreeexplanation{prolactin receptor, \textit{PRLR}}}
\label{fig:prolactin_receptor_evolution_tree}
\end{figure}

\begin{figure}
<<hoxa11.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="HOXA11" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
                 subtree=c("ENSBTAG00000014738","ENSECAG00000013229")
                 )
@ 
\caption{\genetreeexplanation{\textit{HOXA11}}}
\label{fig:hoxa11_evolution_tree}
\end{figure}
